Semi-quantitative petrographic characterization of Bronze Age quartzite tools – an exploratory approach

### 2022-05-13 analysis/paper/paper.Rmd
Restructured paper with subscripts

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The following code chunk defines prepares most of the data that will be needed and reads in the necessary libraries -->

```{r load-packages-and-data}
library(ChemoSpec)
library(data.table)
library(factoextra)
library(ggbiplot)
library(ggplot2)
library(ggh4x)
library(ggsci)
library(grid)
library(gridBase)
library(gridExtra)
library(missMDA)
library(patchwork)
library(prospectr)
library(tidyverse)
library(spectrolab)

#Import descriptive metadata
metadata.csv <-
  read.csv2("./analysis/data/raw_data/metadata_20220119.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))

#Import xrf data, set empty fields to NA
xrf.csv <-
  read.csv2("./analysis/data/raw_data/XRF/xrf_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))

#Import nir data, set empty fields to NA
nir.csv <-
  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

#Import raman data, set empty fields to NA
raman.csv <-
  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220127.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

#Import sample list, set empty fields to NA
samples <-
  read.csv2("./analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na.strings=c("", "NA"))

#prepare color/fill and symbols for score plots
pca.colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
pca.shapes <- c(24,25,23,22,21)
pca.munsell_hue <- c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9")

#If including references
#pca.colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#DF5555", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#BFBFBF", "#DFDFDF", "#FFFFFF")
#pca.munsell_hue <- c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "2.5YR", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N7", "N8", "N9")
```

@@

The last decades have seen an increase in the interest for non-flint lithics and raw material studies in European archaeology, providing much needed development in a previously neglected field [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. In regions where local flint and obsidian sources are non-existent, such as Northern Fennoscandia, other materials have been the primary target for lithic production^[This is not to say that flint was not available or used in these regions] (e.g. quartz, quartzite, slate) [@baudou78; @rankama_etal06; @olausson_etal12]. With the lack of any natural geological source of flint in Northern Sweden both quartz and quartzite are instrumental for the inference of settlement and mobility within the region. While quartz and quartzite technology have been part of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. One reason for the lack of methodological development in North Swedish research could be the complex geological background of the region and the abundance of moraine formations [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The degree to which geological materials have been mixed and transported demands considerable effort in order to reconstruct the events that have formed the modern day landscape. While tentaive interpretations of lithic material sourcing tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12], both quartz and quartzite could potentially have been sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

@@

# Material
```{r select-point-data}
#Filter data to focus on points, point fragments and preforms made from quartz/quartzite materials
sample.list <-
  metadata.csv %>%
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

#Reorder type factor levels
sample.list$type <-
  factor(sample.list$type, levels=c("Preform", "Point fragment", "Point"))

#Reorder parish factor levels
sample.list$parish <- 
  factor(sample.list$parish, levels = c("Bjurholm","Bureå","Burträsk", "Bygdeå", "Degerfors", "Dorotea", "Lycksele", "Norsjö", "Skellefteå", "Sorsele", "Stensele", "Tärna", "Umeå", "Vännäs", "Vilhelmina", "Åsele"))
```

```{r sample-count, results='hide'}
#Filter the sample data to exclude reference samples
tot_samples <- 
  metadata.csv %>% 
    filter(type != "Reference")

#Filter sample data to focus on site Lycksele 301
lycksele <-
  metadata.csv %>% 
    filter(site_id == "Lycksele 301")

#Then filter to focus on quartz/quartzite material that are points and preforms 
lycksele_301 <-
  lycksele %>% 
    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | 
             material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")

#Create vector of unique samples from site Lycksele 301
lycksele_301 <-
unique(lycksele_301$sample_id)

@@

#Create a vector in order count number of unique rows in sample ID column for the number of points/preforms included in analysis
sample_uniq <- 
  unique(sample.list$sample_id)
```

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden. A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz and quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman scattering and XRF. The material features a mainly black, grey or white hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent. The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent). A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89]. Previous research suggests a period of use that falls between ca. 3 800 - 2 000 BP [@kaeck09 p.60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite [@baudou78].

A large portion of the collections stored at North Swedish museums were recovered from excavations and surveys performed in conjunction with the river regulations during the 1940s – 1980s [@biornstad06; @janson_hvarfner60], as well as subsequent follow-up survey projects [e.g. @andersson15]. This has likely had a contributing factor to the distribution pattern of the artefacts observed in fig. \@ref(fig:distribution-map) as the known sites with finds of points and preforms are concentrated around the parishes of Dorotea, Lycksele^[In this case the majority of the samples from Lycksele are from the rock painting site of Lycksele 301], Vilhelmina and Tärnaby. This can be seen more clearly in fig. \@ref(fig:distribution-by-parish) with the four parishes accounting for almost ¾ of the total no. of samples.

Västerbotten museum have come far in their digitization effort, making large portions of their collections searchable online. Based on a query of all points and preforms labelled as *quartz*, *brecciated quartz* or *quartzite* a list was generated from their database. Sampling of this list of artefacts was not random. Sites located in areas with few finds were included first in order to increase the geographical distribution, with sampling of areas with higher site density making up the remaining bulk of the dataset. None of these were from the northernmost Skellefte river, however, and thus a selection of 5 artefacts each from three well-known sites stored at the collections of Skellefte museum were included. This resulted in a total dataset of `r length(tot_uniq)` artefacts, and after documentation and looking up missing information, `r length(sample_uniq)` of these were determined to be either points or preforms made from quartz or quartzite material.

@@

```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of sampled sites within Västerbotten County, Sweden. The distribution of sampled sites is likely affected by the excavation and survey history of the region. White: Sampled sites; Black: Parish centroid for measured samples lacking site coordinates. Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
knitr::include_graphics('../figures/Distribution.jpeg', error = FALSE)
```

@@

```{r distribution-by-parish, fig.width=12, fig.height=10, fig.fullwidth=TRUE, fig.cap= paste("Distribution of the", length(sample_uniq), "points and preforms across the parishes in Västerbotten, Sweden. Note that while Lycksele feature a large number of samples", length(lycksele_301), "of these are from the site of Lycksele 301")}

#Stacked bar chart with counts of samples per parish and type
ggplot(sample.list, aes(x= fct_rev(parish), fill = type)) +
  geom_bar() +
  theme_minimal() +
  labs(y = "Count",
       fill = "Type") +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(face = "bold"),
    axis.title.y=element_blank(),
    axis.text.y = element_text(size = 10, face = "bold", colour = "black"),
    axis.title.x = element_text(size = 12, face = "bold", colour = "black")) +
  scale_fill_jco() +
  coord_flip() +
  geom_text(stat ='count', aes(group = parish, label = ..count..), hjust = -0.5)
```

While some lithics were petrographically analysed in the *Early Norrland* project [@biornstad68; @ahman67] it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these. It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature. In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz and brecciated quartzite^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit*]. Those that had no prior material classification registered were sorted into either the quartz or quartzite category. The reason for including quartz in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.

Another reason for keeping a quartz category is the group referred to as *brecciated quartz*. In the above mentioned petrographic study of the artefacts recovered along the Ume river valley Åhman [-@ahman67 p.8] briefly touches upon the definition of *brecciated quartz*, also referring to it as a *mylonite quartz*. Åhman describes the material as having formed from a quartz vein which has fractured due to movements in the mountain massives and recrystallized via pressure. While the translucent material may give the impression of an unusually pure quartzite, Åhman states that the thin-section analysis reveals the fragmented configuration of quartz. *Brecciated quartz* is a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91 p.24].

@@

Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset [@geladi_linderholm20]. PCA modelling was performed using the base R *prcomp()* function [@Rteam17; @rcoreteam21].   
 \linebreak

@@

The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling. As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL (or "missing values"), during PCA. Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.

@@

## NIR - visual range
Examining the average spectra of the different colour groups (dark, light, white and colourless), as expected, the darker material feature higher absorbance of light than the other groups. No other distinctive features can be seen in the visual band, that could potentially be used to distinguish them. It is, however, apparent that while the more extreme hues of dark and white material can be isolated more clearly based on their spectra, there also exists a wider greyscale of material that does not conform as easily to these broader groups.

The cumulative explained variance for the first three components in the PCA is ~ 99.9%, with the accompanying loadings in fig. \@ref(fig:vis-nir-loadings). Most of the contribution to the first Principal Component (PC) can be found within the red part of the spectrum (~ 600 nm - 750 nm). This can be seen in the score plot where darker and lighter materials are separated along PC 1, as darker samples would feature a higher absorbance within the red part of the spectrum (fig. \@ref(fig:vis-nir-pca)). In contrast, most of the contribution to PC 2 occurs within the blue-violet part of the spectrum, which does not seem to alter the structure of the dataset in any significant way. PC 3 feature high contributions within the blue-green range, which also seems to draw the light and dark samples that feature a blue-gray tint towards the top of the score plot.

While the overall structure of the model seems to group colourless and white materials towards the bottom left, with dark material to the right, there are a number of samples that break this pattern. This suggests that the perceived colour of a material does not necessarily match what is detected within the visible spectrum of light. It is also difficult to separate out the greyscale of colours that occur in between the "black" and "white" groups, where both light gray and semi-translucent samples end up. All in all, the results echo the complexity and unreliability that exists within the topic of petrographic colour.

@@

```{r nir-prep}
#aggregate observations by group(sample) and calculate average of wavelength measurements
nir.averaged <- 
  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)

#merge NIR data with metadata
nir.merged <- 
  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))

#select and filter NIR data to focus on relevant material and artefact types
Points.nir <-
  nir.merged %>%
  filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

#Fill the NA fields in the munsell_hue column and mark them as colourless samples
Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
```

```{r vis-nir-loadings, fig.width=20, fig.height=15, fig.cap="Loading lines for PC 1 - 3. The two first components mainly describes features with a linear increase in absorbance across the visual spectrum, but the 3rd component demonstrates more variation, with lower absorbance within the blue-yellow region. "}

#perform PCA with mean-centering
visnir.pca <-
  prcomp(Points.nir[,c(80:430)], center = TRUE, scale = FALSE)

#Filter NIR data to focus on the material with a dark hue 
#and select the NIR range 1 000 - 2 500 nm
#vis.dark <- Points.nir %>% 
#  filter(hue == "Dark") %>% 
#  select(c(1:28),c(79:429))

#spec.vd <- 
#      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:28)

#vis.light <- Points.nir %>% 
#  filter(hue == "Light") %>% 
#  select(c(1:28),c(79:429))

#spec.vl <- 
#  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:28)

#vis.colourless <- Points.nir %>% 
#  filter(hue == "Colourless") %>% 
#  select(c(1:28),c(79:429))

#spec.vt <- 
#  as_spectra(vis.colourless, name_idx = 1, meta_idxs = 2:28)

#vis.white <- Points.nir %>% 
#  filter(hue == "White") %>% 
#  select(c(1:28),c(79:429))

#spec.vw <- 
#  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:28)

#VISNIR LOADING LINES
#For some reason invisible() has to be used to prevent console output showing up in knitted document
par(mfrow=c(3,1))
invisible(sapply(1:3, function(comp){
  plot(400:750, visnir.pca$rotation[,comp], type='l', lwd=2,
      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(400,750,5))}))

#par(mfrow=c(4,1))

#plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())

#plot(mean(spec.vl), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")

#plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Colourless")

#plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
```

@@

```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plots for the 400 nm – 750 nm band of the NIR data. Top: PC 1 and PC 2, bottom: PC 1 and PC 3. PC 1 seems to group dark-light material on opposite sides of the plots while PC 2 does little to change the strucure. PC 3 seems to position material with a blue-gray tint towards the top of the score plot."}

#prepare a basic score plot with fviz_pca_ind using PC1 and PC2
basic_plot1 <-
  fviz_pca_ind(visnir.pca, axes = c(1,2), label="none")
basic_plot2 <-
  fviz_pca_ind(visnir.pca, axes = c(1,3), label="none")

#prepare labels for PCs
vis.pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=2)
vis.pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=2)
vis.pc3var <- round(summary(visnir.pca)$importance[2,3]*100, digits=2)
vis.pc1lab <- paste0("PC1 (",as.character(vis.pc1var),"%)")
vis.pc2lab <- paste0("PC2 (",as.character(vis.pc2var),"%)")
vis.pc3lab <- paste0("PC3 (",as.character(vis.pc3var),"%)")

#VISNIR LOADING LINES
#par(mfrow=c(3,1))
#sapply(1:3, function(comp){
#  plot(400:750, visnir.pca$rotation[,comp], type='l', lwd=2,
#      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(400,750,6))})

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
visnir.p1 <-
  ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
             aes(x=x, y=y, 
                 shape = material, 
                 fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Near infrared 400 - 750 nm") +
  labs(x = vis.pc1lab,
       y = vis.pc2lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", 
                             title.hjust = 0.5)) +
  theme(legend.position = "none")

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  error = TRUE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300,
  encoding = "UTF-8"
)
```

<!-- The following function can be used to suppress message outputs -->
```{r}	
quietly <- function(x) {	
  sink(tempfile())	
  on.exit(sink())	
  invisible(force(suppressMessages(x)))	
}	
```

@@

Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF). The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery. Although the petrographic characterization of the material mainly establishes a point of geological origin for the object it is still an important first step in reconstructing the *chaîne opératoire*. Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these. The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. Prieto et al. [-@prieto_etal19 p. 15] highlights the current bias that exists in the understanding of raw material management in regions where the local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions]. In northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region. While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. Although flint and slate have featured in discussions of trade and exchange [@halen94 pp. 115-122; @lundberg97 pp. 161-171] within the regon, another reason for the lack of methodological development in North Swedish research could be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape. The tentative interpretations of lithic material sourcing within northern Sweden tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not account for the possibility that the material sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

@@

# Material
```{r sample-count, results='hide'}
suppressPackageStartupMessages(library(tidyverse))

#Import descriptive metadata
metadata.csv <-
  read.csv2("./analysis/data/raw_data/metadata_20220407.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))

#Filter the sample data to exclude reference samples
tot_samples <- 
  metadata.csv %>% 
  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
           site_id == "Åsele 182" |  site_id == "Åsele 188" |site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
         type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite")

@@

bquartz <- unique(tot_samples %>% 
  filter(material == "Brecciated quartz"))$sample_id

quartz <- unique(tot_samples %>% 
  filter(material == "Quartz"))$sample_id

quartzite <- unique(tot_samples %>% 
  filter(material == "Quartzite"))$sample_id
```

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites along the river of Ångermanälven, Västerbotten County, Sweden. The study area ranges from the lake of Kultsjön in the southwest of the County to the town of Åsele in the southeast. A dataset of `r length(tot_uniq)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF. The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent. The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent). A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89]. Previous research suggests a period of use that falls between ca. 3 800 - 2 000 BP [@kaeck09 p.60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite [@baudou78].

A large portion of the collections stored at North Swedish museums were recovered from excavations and surveys performed in conjunction with the river regulations during the 1940s – 1980s [@biornstad06; @janson_hvarfner60], as well as subsequent follow-up survey projects [e.g. @andersson15]. Vilhelmina parish in particular has been the target for numerous excavations and survey projects during this period.

@@

```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden. Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
knitr::include_graphics('../figures/site_distribution.jpeg', error = FALSE)
```

\linebreak
The dataset includes three main types of registered petrographic material: quartz, quartzite, and brecciated quartz (Swe. "*brecciekvarts*"). Those that had no prior classification available were sorted into either the quartz or quartzite category.This distribution of samples into these categories are as follows: `r length(bquartz)` brecciated quartz, `r length(quartz)` quartz, and `r length(quartzite)` quartzite samples. The inclusion of quartz and brecciated quartz in the study serves two purposes. The first being to test the potential for separating the material types based on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification. In his study of the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67 p.8] analysed artefacts from sites located along the Ume river valley. He briefly touches upon the definition of *brecciated quartz* (also referred to as *mylonite quartz*) and describes the material as having formed from a quartz vein which has fractured and then recrystallized via pressure. While the translucent material may give the impression of an unusually pure quartzite, Åhman states that the thin-section analysis reveals the fragmented configuration of quartz. *Brecciated quartz* in particular is a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91 p.24]. As the material frequently has been characterized based on megascopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.

@@

```{r sample-hue-pictures, fig.cap= "An assortment of points demonstrating the variation in color and heterogeneity of the material."}
knitr::include_graphics('../figures/sample_hue.jpg', error = FALSE)
```

@@

Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset [@geladi_linderholm20]. PCA has proven useful for exploring complex spectroscopic datasets [@sciuto19], and identifying clusters and trends that relate to the spectral fingerprints of the material. These groupings can then be tested with supervised methods, and in the construction of classification models. Multivariate analysis methods, such as PCA, are also useful for overcoming some of the issues related to so-called "data closure" [@reimann_etal12]. The compositional data add up to a constant (100%), which is described as the *closed sum* effect. Correlation analysis is directly affected by this *closure* effect and therefore not suitable. With suitable preprocessing the PCA model can reveal the structure and relations between the variables (elements) more accurately. PCA modelling in this case was performed using the base R *prcomp()* function [@Rteam17; @rcoreteam21].     
 \linebreak

@@

The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling. As the resulting dataset also features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.

@@

@@

## NIR - near infrared range
```{r nir-sample-count, results='hide'}
#Filter out the dark quartzite samples that feature peaks of interest in their spectra 
points.d <-
  metadata.csv %>%
  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
           site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
           site_id == "Åsele 99", 
         type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
    filter(!sample_id %in% c("153","167","168","169","172","174","175","177","182","183","190","191","193","194","196","198","200","204","207","210","213","214",
                           "215","216","229","234","235","237","238","251","262","265","268","269","272","278","281","282","359","377","385","392","393","397","405",
                           "406","410","411","413","414","415","416","417","424","425","426","428","430","432","55","56")) %>% 
  filter(hue == "Dark")

#pass to vector so that length() can be used in text to quote nr. of dark samples
spec.d <- 
  points.d$sample_id
```

It is well-known that NIRS is less effective on translucent material, such as a pure quartz crystal, as the abscence of impurities in the material will result in little to no reflection/absorbance of light [@hunt77]. Similarly it can be difficult to work with dark materials as most of the light will be absorbed, thus obscuring any potential spectral features. This effect can be seen in figure \@ref(fig:nir-spectra-summary) which shows a summary of the centered NIR spectra by color group. The dark group generally features less distinct peaks than the material with a lighter hue. This is not to say that all spectra collected on dark quartzite features less information, but any model including them will affect the clustering of the dataset as a whole. Filtering out those spectra that feature no peaks or mainly noise leaves `r length(spec.d)` samples in the dark group. Excluding the noisier spectra, and those with no discernable peaks, there is still a noticeable difference between the dark and light group. By applying 2^nd^ derivative Savitszky-Golay transformation on the dark spectra broad spectral features can be filtered out [@chryssikos_gates pp.78-81]. Comparing the centered spectra to the derivative it becomes clear that there are significant peaks that may be of relevance to the characterization of the material (fig. \@ref(fig:nir-savitzky-transform)).

@@

```{r nir-spectra-summary, fig.cap= "nir spectra summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/001-nir-spectra-summary.R"), echo = FALSE}
quietly(source("./analysis/paper/001-nir-spectra-summary.R", encoding = "UTF-8"))
knitr::include_graphics('../figures/001-nir-spectra-summary.png', error = FALSE)
```

\linebreak
 
```{r nir-spectra-dark, fig.cap= "nir spectra summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/002-nir-dark-filtered-summary.R"), echo = FALSE}
quietly(source("./analysis/paper/002-nir-dark-filtered-summary.R", encoding = "UTF-8"))
knitr::include_graphics('../figures/002-nir-dark-filtered-summary.png', error = FALSE)
```
